Tropical Mathematics for the Lambda Calculus

### 2023-01-17 section6.tex
commit

As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm. This viewpoint gives a metric flavor to $\LREL$, and allowed us to relate differential and metric structure. Yet, how far can this correspondence between tropical linear/non-linear algebra and metric topology be pushed?

For instance, natural questions are: (1) Is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules? (2) Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}. 
This correspondence, recalled in this section, and the fact that it leads to the definition of a linear $\Lawv$-category (i.e.~a model of the linear $\lambda$-calculus) of genealized metric spaces extending $\LREL$, is mostly a \emph{collage} of  \emph{folklore} results with more abstract results contained in recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}.
Beyond this reconstruction, our main contribution is to show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by a suitable generalization of the construction of the exponential comonad of $\LREL$.

@@

A $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$ (notice that this is the usual notion of module over the Lawvere quantale $\Lawv$, not that of module over the tropical semi-ring).
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

@@

 While the $\Lawv^{X}$ have a fixed base, for an arbitrary $\Lawv$-module one can retrieve a base via the \emph{Yoneda embedding}
$\Yon: M \to \Hom(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$.

\begin{proposition}\label{prop:yonemod}
For any $\Lawv$-module $M$, the Yoneda embedding has a left-adjoint $\sup(f)=\bigvee_{x\in M}f(x)\star x$.
\end{proposition}

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a linear $\Lawv$-category.
\end{proposition}
\begin{proof}[Proof sketch]
The hom-sets $\Hom(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices (see e.g.~\cite{Russo2007}), under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$.
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.

Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
Finally, both products and coproducts in $\Mod$ are given by the Cartesian products of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
\end{proof}

\begin{remark}
The structure of linear $\Lawv$-category of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$. Indeed, one can prove that $\Hom(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $ (using the fact that module homomorphisms are in bijection with $\Lawv$-matrices), and 
moreover $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ and 
$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
\end{remark}

@@

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}, or as a \emph{quasi-metric space}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

@@

For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
Notice that, when $X$ has the discrete metric, $[X,\Lawv]$ coincides with $\Lawv^{X}$.
The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.

Interestingly, a fundamental example of $\Lawv$-categories is provided by $\Lawv$-modules.

@@

Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973}. Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{Willerton2013, Rutten1998}).

@@

\begin{definition}
Let $X,Y,Z$ be $\Lawv$-enriched category,
$\Phi: Z\pfun Y$ be a distributor and  $f:Y\to X$ be a functor.
A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in z$ and $x\in X$
\begin{align}
X(g(z),x)= \sup_{y\in Y}\left\{X(f(y),x)\menus \Phi(y,z)\right\}
\end{align} 
A functor $f:X\to Y$ is said \emph{continuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category 
$X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
\end{definition}

@@

A useful alternative characterization of complete $\Lawv$-categories is the following:
\begin{proposition}
A $\Lawv$-category is complete iff $\Yon$ has a left-adjoint. 
\end{proposition}
Indeed, using this fact, together with Proposition \ref{prop:yonemod}, we arrive at the following:
\begin{proposition}
For any $\Lawv$-module, the associated $\Lawv$-category is complete. 
\end{proposition}

@@

\begin{theorem}
$\Mod$ and $\GMet$ are isomorphic categories.
\end{theorem}

Since $\Mod$ (and thus $\GMet$ too) is a linear $\Lawv$-category, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,X_{i},Y$, we have that:
\begin{itemize} 
\item the distance on the hom-set $\Hom(X,Y)$ is given by 
\begin{align}
\Hom(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
\end{align}
\item the distance on the tensor $X\otimes Y$ is given by
\begin{align}
(X\otimes Y)(\alpha, \beta)=
\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
\end{align}
where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
and thus coincides with the sum-metric over basic tensors;
\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
\begin{align}
(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
\end{align}

\end{itemize}

@@

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting the fact that tensors and products commute in $\Mod$, we define $!$ as the product of the symmetric algebras $!_{n}$.

@@

To prove that the map $h:\Sym_{n}(M)\to M^{\otimes_{n}} $ is the equalizer of the diagram formed by all homomorphisms $\langle \sigma\rangle$, it is useful to provide an alternative characterization of it.

\begin{definition}
Let $M$ be a $\Lawv$-module and $n\in \BB N$. An element $x\in M^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
$\langle \sigma \rangle (x)=x$.

 A \emph{$\Lawv$-multiset} (with $n$ elements) is an element of $M^{\otimes_{n}}$ of the form 
 \begin{align}
 [x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}
 x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}
 \end{align}
where $x_{1},\dots, x_{n}\in M$.
\end{definition}

\begin{proposition}
Any $\Lawv$-multiset is p-invariant. Moreover, the set $!_{n}M$ of p-invariant elements of $M^{\otimes_{n}}$ is a $\Lawv$-submodule of $M^{\otimes_{n}}$, in which element can be written as a join of $\Lawv$-multisets.
\end{proposition}

@@

Also in this case of exponentials, the constructions above generalize the situation in $\LREL$:
\begin{proposition}
For any set $X$, 
$!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{n}(X)}$, where
$\C M_{n}(X)$ is the set of multisets of $X$ of cardinality $n$, and 
$!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$. 
\end{proposition}

To conclude, we mush show that $\Mod$ is actually a model of $\STDLC$, that is, a cartesian closed differential category:
\begin{enumerate}
\item $\Mod$ is a left-additive-CCC, since each hom-set is a commutative monoid with respect to $\infty$ and $\min$, and the cartesian closed structure is well-behaved with respect to that;

\item $\Mod$ can be equipped with a differential operator defined, for $f:!M\to N$, by 
\begin{align}\label{eq:dermod}
\Der[f](\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
\end{align}
where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, 
satisfying all axioms (D1)-(D7) and (DCurry), see the Appendix.
\end{enumerate}

This leads then to:
\begin{theorem}
$\Mod_{!}$ (equivalently, $\GMet_{!}$), equipped with $\Der$, is a $CC\partial C$.
\end{theorem}

Notice that, when $f:!(\Lawv^{X})\to \Lawv^{Y}$, so that $f$ is described by a matrix $\widehat f: \multiset(X)\times Y\to \Lawv$, one can check that the matrix associated with \eqref{eq:dermod} coincides with the definition of the differential operator in $\LREL$.

>>>>

As we have seen, the morphisms of $\LREL$ can be seen as continuous functions between the $\Lawv$-modules $\Lawv^{X}$, when the latter are taken with the metric induced by the $\infty$-norm. This viewpoint gives a metric flavor to $\LREL$, and allowed us to relate differential and metric structure. Yet, how far can this correspondence be pushed?
In particular, is this correspondence restricted to $\Lawv$-modules of the form $\Lawv^{X}$ (i.e.~with a fixed base), or does it hold in some sense for arbitrary $\Lawv$-modules? Is this correspondence restricted to the $\infty$-norm metric, or does it hold for other metrics too?

An answer to these questions comes from an elegant categorical correspondence between tropical linear algebra and the theory of \emph{generalized metric spaces}, initiated by Lawvere's pioneering work \cite{Lawvere1973}, and at the heart of the emergent field of \emph{monoidal topology} \cite{Hofmann2014, Stubbe2014}. 
In this section we first reconstruct this correspondence, by combining some {folklore} results with more abstract ones from recent literature in enriched category theory \cite{Fuji, Stubbe2006, Shen2014}. Then we show that this correspondence can be lifted to a model of the full differential $\lambda$-calculus, by suitably generalizing the construction of the exponential comonad of $\LREL$.

@@

A $\Lawv$-module is a triple $(M,\preceq, \star)$ where $(M, \preceq)$ is a sup-lattice, and $\star: \Lawv \times M \to M$ is a continuous (left-)action of $\Lawv$ on it, where continuous means that $\star$ commutes with both joins in $\Lawv$ and in $M$.% (notice that this is indeed the usual notion of module over the Lawvere quantale $\Lawv$).
A $\Lawv$-module homomorphism is a map $f:M\to N$ commuting with both joins and the $\Lawv$-action. We let $\Mod$ indicate the category of $\Lawv$-modules and their homomorphisms.

@@

 While the $\Lawv^{X}$ have a fixed base, for an arbitrary $\Lawv$-module one can retrieve a base via the \emph{Yoneda embedding}
$\Yon: M \to \Mod(M^{\op},\Lawv)$, where $\Yon(x)(y)=\inf\{\epsilon\mid \epsilon\star y\succeq x\}$.

\begin{proposition}[cf.~\cite{Stubbe2014, Shen2014}]\label{prop:yonemod}
For any $\Lawv$-module $M$, the Yoneda embedding has a left-adjoint $\sup(f)=\bigvee_{x\in M}f(x)\star x$.
\end{proposition}

Like $\LREL$, the category $\Mod$ has the relevant structure to interpret the linear $\lambda$-calculus:
\begin{proposition}
$\Mod$ is a SMCC.
\end{proposition}
%\begin{proof}
The hom-sets $\Mod(M,N)$ have a natural $\Lawv$-module structure, defined pointwise. Moreover, the tensor product of $\Lawv$-modules $M\otimes N$ can be defined as the quotient of the usual tensor product of sup-lattices under the smallest congruence containing all pairs $(\{(\epsilon \star x,y)\},\{(x,\epsilon\star y)\})$ (see e.g.~\cite{Russo2007}).
Notably, any element of $M\otimes N$ can be identified with a join of basic tensors $x\otimes y$, corresponding to the equivalence class of the pair $\langle x,y\rangle$.
Beyond the required adjointness of the internal hom and the tensor, one can check that $\Mod$ is actually \emph{$^{*}$-autonomous}, since it satisfies $(M^{\op})^{\op}\simeq M$ and 
$\Hom(M,N^{\op})\simeq (M\otimes N)^{\op}$.
Finally, both products and coproducts in $\Mod$ are given by the Cartesian products of the underlying posets, with action defined pointwise.
%
%, similarly to the case of sup-lattices \cite{}, as the quotient of the free sup-lattice $\C P(M\times N)$ under the smallest congruence containing all pairs $((\bigvee A,y),\bigcup_{a\in A}\{(a,y)\})$, for $A\subseteq M, y\in N$, 
%$((x,\bigvee B),\bigcup_{b\in B}\{(x,b)\})$, for all $x\in A$, $B\subseteq N$, and all pairws $\{
%\end{proof}

\begin{remark}
The SMCC structure of $\Mod$ coincides with that of $\LREL$ for the modules $\Lawv^{X}$: we already know that $\Mod(\Lawv^{X},\Lawv^{Y})\simeq \Lawv^{X\times Y} $, and 
one can prove $\Lawv^{X}\otimes \Lawv^{Y}\simeq \Lawv^{X\times Y} $ (cf.~\cite{Russo2007}).% and 
%$\Pi_{i\in I}\Lawv^{X_{i}}\simeq \Lawv^{\coprod_{i\in I}X_{i}}$.
\end{remark}

@@

Lawvere was the first to observe that a metric space can be described as a \emph{$\Lawv$-enriched} category. Indeed, spelling out the definition, a $\Lawv$-enriched category (in short, a $\Lawv$-category) is given by a set $X$ together with a ``hom-set'' $X(-,-):X\times X\to \Lawv$, satisfying 
\begin{align}
0  & \geq X(x,x) \tag{$\Lawv$-cat 1}\\
X(y,z)+X(x,y)&\geq  X(x,z) \tag{$\Lawv$-cat 2}
\end{align}
This structure is often referred as a \emph{generalized metric space} \cite{Lawvere1973, Hofmann2014, Stubbe2014}.%, or as a \emph{quasi-metric space}. 
Notice that a $\Lawv$-enriched \emph{functor} between $\Lawv$-categories is just a non-expansive map $f:X\to Y$, i.e.~it must satisfy $Y(f(x),f(y))\leq X(x,y)$.

@@

For any $\Lawv$-category $X$, the presheafs $[X^{\op},\Lawv]$ on $X$ form another $\Lawv$-category, with metric $[X^{\op},\Lawv](f,g)= \sup_{x\in X}\Lawv(f(x),g(x))$.
Notice that, when $X$ has the discrete metric, the metric space $[X,\Lawv]^{\sym}$ coincides with $\Lawv^{X}$ with the norm metric.
The \emph{Yoneda embedding} is the faithful functor $\Yon: X\to [X^{\op},\Lawv]$ given by $\Yon(x)(y)=X(y,x)$.

Actually, a well-studied example of $\Lawv$-categories is provided by $\Lawv$-modules.

@@

Lawvere also observed that Cauchy-completeness can be expressed in categorical language as the representability in $X$ of certain presheaves in $[X^{\op},\Lawv]$ \cite{Lawvere1973, Rutten1998}. Category theory suggests a yet stronger notion of completeness, corresponding to the existence of all \emph{weighted colimits} (for a comparison of different notions of completeness on $\Lawv$-categories, see \cite{Willerton2013, Rutten1998}).

@@

\begin{definition}
Let $X,Y,Z$ be $\Lawv$-categories,
$\Phi: Z\pfun Y$ be a distributor and  $f:Y\to X$ be a functor.
A functor $g:Z\to X$ is the \emph{$\Phi$-weighted colimit of $f$ over $X$}, noted $\colim(\Phi,f)$, if for all $z\in Z$ and $x\in X$
\begin{align}
X(g(z),x)= \sup_{y\in Y}\left\{X(f(y),x)\menus \Phi(y,z)\right\}
\end{align} 
A functor $f:X\to Y$ is \emph{continuous} if it commutes with all existing weighted colimits in $X$, i.e.~$f(\colim(\Phi,g))=\colim(\Phi,f\circ g)$. A $\Lawv$-enriched category 
$X$ is said \emph{categorically complete} (or just \emph{complete}) if all weighted colimits over $X$ exist. 
\end{definition}

@@

A useful alternative characterization of complete $\Lawv$-categories is the following:
\begin{proposition}[cf.~\cite{Stubbe2006, Shen2014}]
A $\Lawv$-category is complete iff $\Yon$ has a left-adjoint. 
\end{proposition}
\begin{proof}
For one side, if $X$ is complete, one can define $\sup : [X^{\op},\Lawv]\to X$ as a weighted colimit via $X(\sup x,b)=\sup_{a\in X}
X(a,b)-x_{a}$. Conversely, if a left-adjoint $\sup $ exists, one can define $\mathrm{colim}(\Phi,f):= \sup \Psi$, where $\Psi_{a}=\inf_{b\in X}X(a,f(b))+\Phi_{b}$. 
\end{proof}

Indeed, using this fact, together with Proposition \ref{prop:yonemod}, we arrive at the following:
\begin{proposition}
For any $\Lawv$-module, the associated $\Lawv$-category is complete. 
\end{proposition}

@@

\begin{theorem}[cf.~\cite{Stubbe2006}]
$\Mod$ and $\GMet$ are isomorphic categories.
\end{theorem}

Since $\Mod$ (and thus $\GMet$ too) is a SMCC, it is worth making its metric structure explicit. Given complete $\Lawv$-categories $X,Y$, we have that the distance on the hom-set $\Mod(X,Y)$ is given by 
\begin{align}
\Mod(X,Y)(f,g)= \sup_{x\in X}Y(f(x),g(x));
\end{align}
and the distance on basic tensors 
of $X\otimes Y$ is given by
\begin{align}
%(X\otimes Y)(\alpha, \beta)=
%\sup_{i}\inf_{j}X(x_{i},x'_{j})+Y(y_{j},y'_{j}),
(X\otimes Y)(x\otimes y, x'\otimes y')=
X(x,x')+Y(y,y')
\end{align}
and is extended to arbitrary joins by continuity.
%where $\alpha=\bigvee_{i}x_{i}\otimes y_{i}$ and 
%$\beta=\bigvee_{j}x'_{j}\otimes y'_{j}$, 
%and thus coincides with the sum-metric over basic tensors;
%\item the distance on the bi-product $\Pi_{i\in i}X_{i}$ is given by
%\begin{align}
%(\Pi_{i\in I}X_{i})(f,g)= \sup_{i\in I}X_{i}(f(x),g(x));
%\end{align}

%\end{itemize}

@@

First, we need to define a Lafont exponential $!$ over $\Mod$, and for this we use a well-known recipe from \cite{Mellies2018, Manzo2013}, that is: we first define a symmetric algebra $\Sym_{n}(M)$ as the equalizer of all permutative actions on $n$-tensors $M\otimes \dots \otimes M$; then, exploiting suitable properties of $\Mod$, we may define $!$ as the product of the symmetric algebras.

@@

To prove that the map $[x]\mapsto x:\Sym_{n}(M)\to M^{\otimes_{n}} $ is the equalizer of the diagram formed by all morphisms $\langle \sigma\rangle$, it is useful to provide an alternative characterization of it.

\begin{definition}
Let $M$ be a $\Lawv$-module and $n\in \BB N$. An element $x\in M^{\otimes_{n}}$ is said \emph{permutation-invariant} (in short, \emph{p-invariant}) if for all $\sigma\in \F S_{n}$, 
$\langle \sigma \rangle (x)=x$. 
 A \emph{$\Lawv$-multiset} (with $n$ elements) is an element of $M^{\otimes_{n}}$ of the form 
$ [x_{1},\dots, x_{n}]:= \bigvee_{\sigma\in \F S_{n}}
 x_{\sigma(1)}\otimes \dots \otimes x_{\sigma(n)}$, where $x_{1},\dots, x_{n}\in M$.
\end{definition}

\begin{proposition}
Any $\Lawv$-multiset is p-invariant. Moreover, the set $!_{n}M$ of p-invariant elements of $M^{\otimes_{n}}$ is a $\Lawv$-submodule of $M^{\otimes_{n}}$, whose elements can be written as joins of $\Lawv$-multisets.
\end{proposition}

@@

Also in the case of exponentials, the constructions for $\Mod$ generalize those of $\LREL$:
\begin{proposition}
$!_{n}(\Lawv^{X})\simeq \Lawv^{\C M_{\leq n}(X)}$, $!(\Lawv^{X})\simeq \Lawv^{\multiset(X)}$. In particular, $\Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$.
\end{proposition}

To conclude, we mush show that $\Mod$ is a CC$\partial$C.
Firstly, $\Mod$ is a left-additive-CCC, since each hom-set is a commutative monoid with respect to $\infty$ and $\min$, and the cartesian closed structure is well-behaved with respect to that.

Secondly, $\Mod$ can be equipped with a differential operator $E$ defined, for $f:!M\to N$, by 
\begin{align}\label{eq:dermod}
Ef(\alpha)=
\bigvee\left\{
f(\beta\cup [x]) \ \Big \vert  \ 
\iota_{n}(\beta)\otimes \iota_{1}(x) \leq S(\alpha)
\right\}
\end{align}
where $\iota_{k}: M_{k}\to \prod_{i\in I}M_{i}$ is the injection morphism given by $\iota_{k}(x)( k)=x$ and $\iota_{k}(x)(i\neq k)=\infty$,
and $S: !(M\times N)\to !M\otimes !N$ is the Seely isomorphism \cite{Mellies2018}, and $E$ satisfies all required axioms.
%\end{enumerate}

Notice that, when $f\in \Mod_{!}(\Lawv^{X},\Lawv^{Y})\simeq \LREL_{!}(X,Y)$, its derivative $E f$ coincides with the derivative $D_{!}f$ described in Section \ref{section4}.

All this leads then to:
\begin{theorem}
$\Mod_{!}$ (equivalently, $\GMet_{!}$), equipped with $E$, is a $CC\partial C$.
\end{theorem}

### 2023-01-17 section8.tex
commit

We have shown that:

- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)

- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)

- this relation \emph{seems} to provide applications in different fields (section 5).

Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
For instance, the general questions are:

- Can we improve the results of section 4 ?

- Can we develop and make the applications of section 5 useful ?

- What does the general setting of section 6 give in terms of theoretical and applied results ?

- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)

- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
Say what could they do and why.

>>>>

In our opinion, the main goals of this paper are two. Firstly,  to
demonstrate the existence of a conceptual bridge between two different well-studied quantitative approaches to higher-order programs, and to highlight the possibility of transferring results and techniques from one approach to the other. 
Secondly, to suggest that tropical mathematics, a
field which has been largely and successfully applied in computer science, could be used to study quantitative properties of higher-order programs.

While the first goal has been developed in detail, and at different levels of abstraction, for the second goal we only sketched a few interesting directions (best-case analysis, log-probabilities, differential privacy). We believe that exploring in more depth these ideas could be a fruitful direction; moreover, 
since both generalized metrics and quantale-modules have been largely applied in computer science, 
a natural question is whether the generalized approach of Section \ref{section6} could  lead to new applications of metric and tropical methods to the $\lambda$-calculus.

Finally, as suggested in Section \ref{section5}, an interesting direction to look at is the semantics of finiteness spaces \cite{Ehrhard2005}, which allows to model typable terms by means of \emph{finitary} Taylor expansions. Indeed, knowing that the application of a program $M$ to $N$ expands as a finite sum of linear applications (hence by a Lipschitz function in our setting), may allow one to predict how sensitive $M$ will be ``around $N$''.
%
%We have shown that:
%
%- intuitively, the interplay between tropical mathematics and $\lam$-calculus \emph{could} relate the metric and differential approaches on approximations of $\lam$-calc (introduction+section 2)
%
%- this relation \emph{can} take place, because the natural category $\LREL$ and its generalised versions are metric models of the differential $\lam$-calculus (section 3 and 6)
%
%- this relation \emph{seems} to provide applications in different fields (section 5).
%
%Therefore, we mainly set the basis for future interplays between all these areas, hopefully motivating the interest in such an interplay.
%For instance, the general questions are:
%
%- Can we improve the results of section 4 ?
%
%- Can we develop and make the applications of section 5 useful ?
%
%- What does the general setting of section 6 give in terms of theoretical and applied results ?
%
%- Do tools from tropical \emph{geometry} provide something for $\lam$-calculus ? (for instance, the role of tropical roots, tropical varieties,...)
%
%- Finally, there is a last point which we think of interest and we did not mention through the paper: the inclusion of finiteness spaces in the picture.
%Say what could they do and why.
